---
title: "spongEffects"
author: "R. Ellenrieder"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(SPONGE)
library(doParallel)
library(foreach)
library(dplyr)

# Register your backend here
num.of.cores <- 1
cl <- makeCluster(num.of.cores) 
registerDoParallel(cl)
```

# Load the data

spatial expression object
```{r}
library(anndata)
adata <- read_h5ad("../datasets/GSM6592055_M8.h5ad")
```

extract expression matrix
```{r}
# Expression matrix (may be sparse)
expr <- adata$X

# Convert to dense matrix if needed
if (inherits(expr, "dgRMatrix")) {
  expr <- as.matrix(expr)
}

# Preview
expr[1:5, 1:8]
```


```{r}
library(data.table)

ceRNA_interactions <- fread("../networks/SPONGE/breast_invasive_carcinoma/breast_invasive_carcinoma_interactionNetwork.csv")

head(ceRNA_interactions)
```

```{r}
ceRNA_centralities <- fread("../networks/SPONGE/breast_invasive_carcinoma/breast_invasive_carcinoma_networkAnalysis.csv")

head(ceRNA_centralities)
```

```{r}
dim(ceRNA_centralities)
length(unique(c(ceRNA_interactions$geneA, ceRNA_interactions$geneB)))
```

filter ceRNA network and add weighted centrality measures information
```{r}
summary(ceRNA_interactions$mscor)
summary(ceRNA_interactions$p.adj)
```

```{r}
filtered_network_centralities <- filter_ceRNA_network(sponge_effects = ceRNA_interactions, Node_Centrality = ceRNA_centralities, add_weighted_centrality=T, mscor.threshold = 0.01, padj.threshold = 0.1)
```

```{r}
dim(ceRNA_interactions)
dim(ceRNA_centralities)

dim(filtered_network_centralities$Sponge.filtered)
dim(filtered_network_centralities$Node_Centrality)
```

```{r}
top_genes <- filtered_network_centralities$Node_Centrality[order(-eigenvector)][,gene][1:100]
```


Discover modules
```{r}
central_gene_modules<-get_central_modules(central_nodes = top_genes, node_centrality = filtered_network_centralities$Node_Centrality)

central_gene_modules
setdiff(union(central_gene_modules$gene, top_genes), intersect(central_gene_modules$gene, top_genes))
```

Define SPONGE modules

```{r}
Sponge.modules <- define_modules(network = filtered_network_centralities$Sponge.filtered, central.modules = central_gene_modules, remove.central = F, set.parallel = F)
# Module size distribution
Size.modules <- sapply(Sponge.modules, length)
```

```{r}
Size.modules
hist(Size.modules)
```

Calculate Enrichment Scores

Overall Enrichment (OE)
single-sample Gene Set Enrichment Analysis (ssGSEA)
Gene Set Variation Analysis (GSVA)

```{r}
class(expr)
str(expr)
```

```{r}
train.modules <- enrichment_modules(Expr.matrix = expr, modules = Sponge.modules, bin.size = 10, min.size = 1, max.size = 2000, min.expr = 1, method = "OE", cores=1)
```

```{r}
t_expr <- t(expr)
```

Machine Learning spongEffects scores













